Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-24 articles/cycle/alps-cycle.tex
Remove text on self-sustained ice domes.

    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
    glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
    and 1\,km. Climate forcing is derived using present-day climate data from
    WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
    offsets from multiple palaeo-climate proxies, among which only the EPICA ice
    core record yields glaciation during marine oxygen isotope stages~4
    (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
    the geological reconstructions, while the other records used result in
    excessive early glacial cycle ice cover and a late Last Glacial Maximum.
    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that Alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences and
    self-sustained ice domes. Modelled maximum ice surface is 861\,m
    higher than observed trimline elevations in the upper Rhone Valley, yet
    our simulation predicts little erosion at high elevation due to cold-based
    ice. Finally, despite the uniform climate forcing, differences in glacier
    catchment hypsometry cause the Last Glacial Maximum advance to be modelled
    as a time-transgressive event, with some glaciers reaching their maximum as
    early as 27\,ka, and others as late as 21\,ka.

@@

    \begin{figure*}[t]
      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
      \caption{%
        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
        (200\,m contours), and ice surface velocity (blue) in the Alps
        24.57\,ka before present, corresponding to the maximum modelled ice
        cover. The modelled LGM ice extent (dashed orange line) and
        geomorphological reconstruction \citep[solid red
        line,][]{Ehlers.etal.2011}, modelled self-sustained ice domes
        (triangles), and some major transfluences (crosses) are shown.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model \citep[black
        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
        last glacial cycle (120--0\,ka), expressed in meters of sea level
        equivalent (m~s.l.e., blue curve). Shaded gray areas indicate the
        timing for Marine Oxygen Isotope Stage (MIS)~2 and MIS~4 according to
        a~global compilation of benthic \chem{\delta^{18}O} records
        \citep{Lisiecki.Raymo.2005}.
        \todo{Thicken the 200\,m contours on all figures. Remove ice domes and
              replace Straniger.}}
      \label{fig:lgmvel}
    \end{figure*}

@@

    Finally, self-sustained ice domes characteristic of ice caps are modelled
    in two locations over Flüela Pass and Ötztal (Fig.~\ref{fig:lgmvel}a).
    These modelled ice domes exhibit radial ice surface slopes and radial flow
    velocities crosscutting the underlying topographic troughs and ridges.
    On the one hand, ice-dispersal centres previously labelled as ice domes
    in the upper Rhone and Rhine Valley \citep{Florineth.Schluchter.1998,
    Kelly.etal.2004, Bini.etal.2009} feature as mere ice divides in our
    simulation. On the other hand, the modelled ice dome over Flüela Pass
    overlaps with the previously mapped Engadine ice dome
    \citep{Florineth.1998}, although its centre is modelled ca.~20\,km north of
    that indicated by geomorphology \citep[Fig.~3]{Florineth.1998}.

@@

    \begin{itemize}
      \item Ice cover is generally underestimated in the north-western Alps and
            overestimated in the eastern and south-western Alps, indicating
            that east-west
            gradients in temperature or precipitation change, absent from our
            model forcing, probably controlled the LGM extent of ice cover in
            the Alps. The observed asymmetric extent of ice north and south of
            the Alps can be explained by the modelled transient nature of the
            LGM extent without involving north-south gradients in temperature
            and precipitation change (Sect.~\ref{sec:extent}).
      \item The LGM ice flow pattern in the Alps was largely controlled by
            subglacial topography, but several transfluences and two
            self-sustained ice domes may have occurred (Sect.~\ref{sec:flow}).
      \item The LGM (maximum) extent was a transient stage in which glaciers
            were out of balance with the
            contemporary climate. Its timing potentially varied across the
            range due to inherent glacier dynamics (Sect.~\ref{sec:timing}).
      \item Ice thickness during the LGM is modelled to be much larger than in
            reconstructions. In average, modelled surface elevation is 861\,m
            above the Rhone Glacier trimlines, which may instead indicate an
            englacial thermal boundary (Sect.~\ref{sec:thickness}).
      \item Alpine glaciers were very dynamic. They quickly
            responded to climate fluctuations and some potentially advanced
            many times over the foreland during the last glacial cycle
            (Sect.~\ref{sec:glaciations}).
    \end{itemize}

>>>>

    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
    glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
    and 1\,km. Climate forcing is derived using present-day climate data from
    WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
    offsets from multiple palaeo-climate proxies, among which only the EPICA ice
    core record yields glaciation during marine oxygen isotope stages~4
    (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
    the geological reconstructions, while the other records used result in
    excessive early glacial cycle ice cover and a late Last Glacial Maximum.
    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that Alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences through the
    mountain passes. Modelled maximum ice surface is 861\,m
    higher than observed trimline elevations in the upper Rhone Valley, yet
    our simulation predicts little erosion at high elevation due to cold-based
    ice. Finally, despite the uniform climate forcing, differences in glacier
    catchment hypsometry cause the Last Glacial Maximum advance to be modelled
    as a time-transgressive event, with some glaciers reaching their maximum as
    early as 27\,ka, and others as late as 21\,ka.

@@

    \begin{figure*}[t]
      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
      \caption{%
        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
        (200\,m contours), and ice surface velocity (blue) in the Alps
        24.57\,ka before present, corresponding to the maximum modelled ice
        cover. The modelled LGM ice extent (dashed orange line) and
        geomorphological reconstruction \citep[solid red
        line,][]{Ehlers.etal.2011}, and some major transfluences across
        mountain passes (crosses) are shown.
        \textbf{(b)} Temperature offset time-series from the EPICA ice core
        used as palaeo-climate forcing for the ice flow model \citep[black
        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
        last glacial cycle (120--0\,ka), expressed in meters of sea level
        equivalent (m~s.l.e., blue curve). Shaded gray areas indicate the
        timing for Marine Oxygen Isotope Stage (MIS)~2 and MIS~4 according to
        a~global compilation of benthic \chem{\delta^{18}O} records
        \citep{Lisiecki.Raymo.2005}.
        \todo{Thicken the 200\,m contours on all figures. Remove ice domes and
              replace Straniger.}}
      \label{fig:lgmvel}
    \end{figure*}

@@

@@

    \begin{itemize}
      \item Ice cover is generally underestimated in the north-western Alps and
            overestimated in the eastern and south-western Alps, indicating
            that east-west
            gradients in temperature or precipitation change, absent from our
            model forcing, probably controlled the LGM extent of ice cover in
            the Alps. The observed asymmetric extent of ice north and south of
            the Alps can be explained by the modelled transient nature of the
            LGM extent without involving north-south gradients in temperature
            and precipitation change (Sect.~\ref{sec:extent}).
      \item The LGM ice flow pattern in the Alps was largely controlled by
            subglacial topography, but transfluences across several mountain
            passes may have occurred (Sect.~\ref{sec:flow}).
      \item The LGM (maximum) extent was a transient stage in which glaciers
            were out of balance with the
            contemporary climate. Its timing potentially varied across the
            range due to inherent glacier dynamics (Sect.~\ref{sec:timing}).
      \item Ice thickness during the LGM is modelled to be much larger than in
            reconstructions. In average, modelled surface elevation is 861\,m
            above the Rhone Glacier trimlines, which may instead indicate an
            englacial thermal boundary (Sect.~\ref{sec:thickness}).
      \item Alpine glaciers were very dynamic. They quickly
            responded to climate fluctuations and some potentially advanced
            many times over the foreland during the last glacial cycle
            (Sect.~\ref{sec:glaciations}).
    \end{itemize}
